---
title: "Can the number of cat adoptions predict the volume of traffic?"
output: html_document
editor_options: 
  chunk_output_type: inline
---

_This is a short introductory example of using R that demonstrates simple processes for cleaning and visualizing data and basic statistical analysis. This report was generated using R Markdown Notebook. In no way should the objective of this example be taken seriously, nor do they reflect the views of the instructors or organization sponsors of this workshop._

# Objective
The feline species are found everywhere in urban areas. Although the vast population of felines do not have "thumbs" to manipulate a steering wheel, increasing research and developement of autonomous vehicles might render their anatomical barrier moot. In preparation of the eventual takeover of autonomous vehicles do we see their infilitration to human households increasing? If so, we first must investigate if there's any relationship between the number of cat adoptions with volume of traffic.

# Methods & Results
## Data sources
Monthly traffic volume data were downloaded from the PORTAL website (https://portal.its.pdx.edu) for the Alberta to NB I-5 station for all years available.

Yearly animal services trends are reported by Multnomah County Animal Services (MCAS) (https://multcopets.org/reports-and-agency-statistics). All reporting by MCAS is presented in a non machine readable format and has been converted into a csv file prior to use for this example. Data reported by MCAS is done once per fiscal year which ranges from July through June.

## Data cleanup
A fair amount of data cleanup happens prior to any analysis; however, with R and packages like Tidyverse make the process of data wrangling less painful. The process of cleaning and tidying data is recorded in each line of code that is written. This makes it easier and faster to catch and fix mistakes for large and small datasets. Below is an example of working with a fairly tidy dataset.
```{r show libraries, eval = F}
# Load libraries needed for tidying data.
library(tidyverse)
library(lubridate)
```
```{r run libraries, include = F, echo = F}
# Load libraries needed for tidying data.
library(tidyverse)
library(lubridate)
```

```{r read in data}
# Read in data files.
alberta_raw <- read.csv("data/I5-NB-Alberta.csv", stringsAsFactors = F, strip.white = T)
cats_raw <- read.csv("data/cats.csv", stringsAsFactors = F, strip.white = T)
```
```{r examine traffic data}
# Examine traffic data structure.
str(alberta_raw)

# Delete rows of incomplete data
alberta <- alberta_raw[complete.cases(alberta_raw),]

# Convert "starttime" into class POSIXct
alberta$starttime <- mdy_hm(alberta$starttime, tz = "America/Los_Angeles")

# Check for completeness of data
alberta_year <- alberta %>%
  mutate(year = year(starttime))
table(alberta_year$year)
alberta_fy <- alberta_year %>%
  filter(starttime > "2005-06-30" & starttime < "2015-07-01")

# Creative way of aggregating volume by fiscal year, July through June
alberta_fy$fy.adjust <- seq(ymd("2005-01-01"), ymd("2014-12-31"), by = "month")
alberta_fy <- alberta_fy %>%
  group_by(Fiscal.Year = year(fy.adjust)) %>%
  summarise(FY.Volume = sum(volume))
alberta_fy
```
```{r examine cat data}
# Examine cat data
str(cats_raw)

# Filter cat data for available years corresponding to traffic data.
cats <- cats_raw %>%
  filter(between(Fiscal.Year, 2005, 2014))
cats
```

## Data visualization
There are a number of visualization packages for R that allow the user to creating and customizing simple static figures to creating and customizing interactive maps and applications. Below are some basic examples of generating plots to examine the cat and traffic data.

```{r datatable of combined sources}
alberta_cats <- left_join(alberta_fy, cats, by = "Fiscal.Year")

alberta_cats_fig <- alberta_cats %>%
  ggplot(aes(x = Adopt, y = FY.Volume))+
  geom_point()
plot(alberta_cats_fig)
```

## Statistics

# Conclusions

Last updated: 2018-07-23  
Back to [Introduction](https://github.com/TC-piRatecat-2018/Introduction)